---
title: "Alternatives and benefits (R)"
---

```{r}
#| include: false
library(hecfdar)
```

## The question

Is a flood-risk project worth building? HEC-FDA answers with **equivalent annual damage (EqAD)
reduced**: compute EqAD without the project, compute it again with the project, and subtract.
This example builds both sides from the ground up -- a scenario, then an alternative, then the
benefits comparison -- chaining `scenario_results()`, `alternative_ead()`, and
`alternative_comparison()` exactly the way a real HEC-FDA study does.

## The annualization math alone

Before chaining the whole pipeline, look at the annualization step by itself. `eqad()` takes an
EAD at a base year and an EAD at a future year and folds them into one equivalent annual number:

```{r}
eqad(35000, 2023, 50000, 2072, 50, 0.07)
```

That reproduces the upstream `ComputeEEAD_Test` oracle, `38835.31`. Three steps produce it:

1. **Interpolate** -- linearly interpolate EAD between the base year and the most likely future
   year, one value per year of the analysis period. If the future year is reached before the
   period ends, every later year holds flat at the future value.
2. **Discount** -- convert each year's EAD to present value at the given discount rate.
3. **Annualize (PVIFA)** -- divide the summed present value by the present-value-interest-factor
   of an annuity over the period, turning a lump sum back into one equivalent annual payment.

Here the base year is 2023, the future year is 2072, and the period of analysis is exactly 50
years -- so the future year lands on the last year of the period and every year gets interpolated,
with no flat tail:

```{r}
base_value = 35000; base_year = 2023
future_value = 50000; future_year = 2072
period = 50; rate = 0.07

years = base_year + 0:(period - 1)
interpolated = base_value + (future_value - base_value) * (0:(period - 1)) / (future_year - base_year)
discounted = interpolated / (1 + rate)^(1:period)
```

```{r}
#| fig-cap: "EAD interpolated year by year (2023-2072) and discounted to present value"
plot(years, interpolated, type = "l", col = "#6b705c", lwd = 2,
     ylim = range(c(interpolated, discounted)), xlab = "year", ylab = "EAD ($)")
lines(years, discounted, col = "#cb997e", lwd = 2)
legend("topleft", legend = c("interpolated EAD", "discounted to present value"),
       col = c("#6b705c", "#cb997e"), lwd = 2, bty = "n")
```

Summing the discounted column and dividing by the annuity factor reproduces the same `38835.31`
`eqad()` returned above:

```{r}
pvifa = (1 - 1 / (1 + rate)^period) / rate
sum(discounted) / pvifa
```

## Without-project alternative

Now the full pipeline. A HEC-FDA alternative needs a base-year and a future-year scenario, each
computed from an impact-area simulation the same way the EAD-simulation example did. This
construct is the upstream `AlternativeResults_Test` case: a Uniform(0, 100000) flow frequency, a
linear flow-stage curve from (0, 0) to (100000, 30), and a 3-point stage-damage curve. The
future-year damages are exactly double the base-year damages -- the project has done nothing yet,
so this is the without-project condition.

```{r}
without_base_sim = list(
  impact_area_id = 1,
  flow_frequency = list(dist = "Uniform", params = c(0, 100000, 1000)),
  flow_stage = list(
    x = c(0, 100000), dist = "Uniform",
    params = list(c(0, 0, 10), c(0, 30, 10))
  ),
  stage_damage = list(list(
    x = c(0, 15, 30), dist = "Uniform",
    params = list(c(0, 0, 10), c(0, 600000, 1), c(0, 600000, 1)),
    damage_category = "residential", asset_category = "structure"
  ))
)
without_future_sim = without_base_sim
without_future_sim$stage_damage[[1]]$params =
  list(c(0, 0, 10), c(0, 1200000, 10), c(0, 1200000, 10))
```

Run each year through `scenario_results()`, then annualize both into one alternative with
`alternative_ead()`:

```{r}
without_base = scenario_results(
  list(without_base_sim), min_iterations = 1L, max_iterations = 1L, deterministic = TRUE
)
without_future = scenario_results(
  list(without_future_sim), min_iterations = 1L, max_iterations = 1L, deterministic = TRUE
)

without = alternative_ead(
  without_base, without_future,
  base_year = 2023, future_year = 2072, period_of_analysis = 50, discount_rate = 0.0275,
  alternative_id = 1L
)
without$mean_eqad
without$base_year_ead
without$future_year_ead
```

Mean EqAD comes out to `208213.80`, the `AlternativeResults_Test` oracle for this exact
construct (period 50, base year 2023, future year 2072, discount rate 2.75%), with a base-year EAD
near 150000 and a future-year EAD near 300000 -- twice as much damage once the future-year curve
takes over.

## With-project alternative

A project reduces damage. Here that project is a levee: 0% failure probability below stage 7.4999,
100% failure probability at 7.5 and above, sitting at a `top_of_levee_elevation` of 7.5. Everything
else -- the flow frequency, the flow-stage curve, the stage-damage curve, the base/future years,
discount rate, and period -- stays exactly the same as the without-project case:

```{r}
levee_spec = list(
  x = c(0, 7.4999, 7.5), dist = "Deterministic", params = list(0, 0, 1),
  damage_category = "residential", asset_category = "structure",
  top_of_levee_elevation = 7.5
)

with_base_sim = c(without_base_sim, list(levee = levee_spec))
with_future_sim = c(without_future_sim, list(levee = levee_spec))

with_base = scenario_results(
  list(with_base_sim), min_iterations = 1L, max_iterations = 1L, deterministic = TRUE
)
with_future = scenario_results(
  list(with_future_sim), min_iterations = 1L, max_iterations = 1L, deterministic = TRUE
)

with_project = alternative_ead(
  with_base, with_future,
  base_year = 2023, future_year = 2072, period_of_analysis = 50, discount_rate = 0.0275,
  alternative_id = 2L
)
with_project$mean_eqad
```

Mean EqAD drops to `161365.70` -- lower than the without-project `208213.80`, because the levee
holds back damage below stage 7.5 in both the base and future year.

## Benefits

The benefit of the project is the damage the levee prevents. `alternative_comparison()` subtracts
the with-project alternative's damage distributions from the without-project alternative's and
reports the reduction:

```{r}
benefits = alternative_comparison(without, with_project)
benefits$reduced
```

The base-year EAD is reduced by `33750` and the future-year EAD by `67500` -- these are the
`AlternativeComparisonReportShould`-style oracle values for this construct, and they are exactly
what a levee holding back stages below 7.5 should prevent: a fixed share of both the base and
future damage, scaled by how much of each curve sits under the levee's protection.

## Single-use handles and validation

`scenario_results()` and `alternative_ead()` both return a `handle` alongside their summary values.
Each handle is **single-use**: `alternative_ead()` takes ownership of the scenario handles it is
given, and `alternative_comparison()` takes ownership of the alternative handles it is given. A
handle already consumed cannot be reused -- recompute the scenario or alternative if you need it
again. That is why this example computes `without_base`/`without_future` and
`with_base`/`with_future` as four separate scenario calls rather than trying to share one.

None of the numbers above are made up. `38835.31`, `208213.80`, `161365.70`, `33750`, and `67500`
are the same literals the dotnet oracle gate (`make oracles`) checks against the real
`HEC.FDA.Model.Alternative`/`AlternativeComparisonReport`, and this example chains
`Scenario.Compute -> Alternative.AnnualizationCompute ->
AlternativeComparisonReport.ComputeAlternativeComparisonReport` exactly the way the upstream
`AlternativeComparisonReportShould`-style tests do -- the full stage-frequency -> stage-damage ->
EAD -> EqAD -> benefits pipeline, validated end to end against the real HEC-FDA.
