---
title: "Stage-damage curves (R)"
---

```{r}
#| include: false
library(hecfdar)
```

## The problem

A stage-damage curve answers one question: for a given flood stage, how much dollar damage does
a group of structures take? HEC-FDA builds that curve from three inputs -- a structure
inventory, a depth-percent-damage function per occupancy type, and the hydraulics that convert a
flood event into a water-surface elevation at each structure -- and adds a stage-frequency curve
so the same compute can report damage against exceedance probability, not just stage.

This example rebuilds the upstream `TractableStageDamageTests` scenario: a simple two-structure
Residential inventory where the answer is checkable by hand. It reproduces the published damage
values for Residential/Structure at stages 12 through 19:

```
{0, 0, 30, 60, 90, 120, 150, 180}
```

## The inventory

Two residential structures. Structure values are 100 and 200; first-floor elevations are 14 and
15; both sit on ground at elevation 12.

```{r}
structures = data.frame(
  fid = c("1", "2"),
  first_floor_elevation = c(14, 15),
  value_structure = c(100, 200),
  damage_category = "Residential",
  occupancy_type = "Residential",
  ground_elevation = 12
)
structures
```

## The occupancy type

An occupancy type carries one depth-percent-damage curve per asset category. Here both curves
are deterministic (no sampling uncertainty): structure damage rises linearly from 0% at depth 0
to 100% at depth 10, and content damage follows a similar rise scaled down slightly. A
content-to-structure value ratio (CSVR) of 50 converts the content percentage into dollars from
the same structure value.

```{r}
depths = 0:10
struct_pct = seq(0, 100, by = 10)
content_pct = c(0, 5, 15, 25, 35, 45, 55, 65, 75, 85, 95)

occupancy_types = list(list(
  name = "Residential", damage_category = "Residential",
  structure_curve = list(x = depths, dist = "Deterministic", params = as.list(struct_pct)),
  content_curve = list(x = depths, dist = "Deterministic", params = as.list(content_pct)),
  content_to_structure_value_ratio = list(central = 50)
))
```

```{r}
#| fig-cap: "Residential depth-percent-damage curves"
plot(depths, struct_pct, type = "l", col = "#6b705c", lwd = 2, ylim = c(0, 100),
     xlab = "depth (ft)", ylab = "damage (%)")
lines(depths, content_pct, col = "#cb997e", lwd = 2)
legend("topleft", legend = c("structure", "content"), col = c("#6b705c", "#cb997e"),
       lwd = 2, bty = "n")
```

## Hydraulics and frequency

Eight exceedance probabilities drive the compute, from the common 0.5 event down to the rare
0.002 event. Each probability has a water-surface-elevation profile, one value per structure;
here the profiles start at stages 13 and 14 for the two structures and rise by 1 ft per
probability step. A graphical stage-frequency curve (stages 12 through 19, equivalent record
length 50) lets the same compute report damage against frequency.

```{r}
probabilities = c(.5, .2, .1, .04, .02, .01, .004, .002)
wses = Reduce(\(prev, i) prev + 1, seq_len(7), init = c(13, 14), accumulate = TRUE)

hydraulics = list(probabilities = probabilities, wses = wses)
stage_frequency = list(probabilities = probabilities, stages = 12:19, erl = 50)
```

## Compute and compare

```{r}
res = stage_damage(
  structures, occupancy_types, hydraulics, stage_frequency,
  stages = 12:19, impact_area_id = 34
)
structure_damage = res[res$damage_category == "Residential" & res$asset_category == "Structure", ]
structure_damage
```

```{r}
#| fig-cap: "Residential/Structure stage-damage curve"
plot(structure_damage$stage, structure_damage$damage, type = "b", col = "#588157", lwd = 2,
     pch = 16, xlab = "stage (ft)", ylab = "damage ($)")
```

The upstream oracle for Residential/Structure at stages 12-19 is `{0, 0, 30, 60, 90, 120, 150,
180}`. The computed values line up exactly:

```{r}
data.frame(
  stage = structure_damage$stage,
  computed = structure_damage$damage,
  oracle = c(0, 0, 30, 60, 90, 120, 150, 180)
)
```

## Why the numbers are trustworthy

This scenario is not a made-up smoke test. It is the exact `TractableStageDamageTests` case from
the upstream HEC-FDA C# source, and the dotnet oracle gate (`make oracles`) replays it against
the real `HEC.FDA.Model.ImpactAreaStageDamage.Compute` and checks that this port's output
reproduces it to tolerance. The `Compute` call above runs the same convergence loop as production
HEC-FDA; on this deterministic scenario it converges after the first chunk, but the loop, the
consequence binning, and the curve construction are the same code the seeded Monte Carlo examples
use.
