---
title: "Distributions (R)"
---

```{r}
#| include: false
library(hecfdar)
```

hecfda's C++ core ports all 13 distribution families from `HEC.FDA.Statistics`. Every family is
addressed the same way: a name string and a numeric parameter vector, in the order the core
factory expects (the same order the `fixtures/distributions/*.json` oracle files use). One
generic set of functions covers all of them: `dist_pdf()`, `dist_cdf()`, `dist_quantile()`, and
`dist_sample()`.

Nine families are reachable this way, through the generic factory dispatch:

| Family | Name string | Parameter order | Notes |
|---|---|---|---|
| Normal | `"Normal"` | `mean, sd, sample_size` | |
| Uniform | `"Uniform"` | `min, max, sample_size` | |
| Triangular | `"Triangular"` | `min, most_likely, max, sample_size` | |
| Deterministic | `"Deterministic"` | `value` | a point mass |
| LogNormal | `"LogNormal"` | `mean, sd, sample_size` | moments of the natural-log-scale data |
| TruncatedNormal | `"TruncatedNormal"` | `mean, sd, min, max, sample_size` | |
| TruncatedLogNormal | `"TruncatedLogNormal"` | `mean, sd, min, max, sample_size` | moments of the natural-log-scale data |
| LogPearsonIII | `"LogPearsonIII"` | `mean, sd, skew, sample_size` | moments of the log10-scale data; the flood-frequency workhorse |
| TruncatedLogPearson3 | `"TruncatedLogPearson3"` | `mean, sd, skew, min, max, sample_size` | moments of the log10-scale data |

Four of the 13 families are not on this list, and calling `dist_pdf()` with any of their names
raises an error rather than constructing one. `Gamma`, `ShiftedGamma`, and `PearsonIII` are
internal helper classes in the C++ core, not members of the `IDistribution` factory, so they have
no name string at all. `Empirical` is a full `IDistribution` member, but is not constructible by
name through the factory: it is built from data (a histogram or a set of quantiles), not from a
short parameter vector, so it has its own construction path rather than a `dist_pdf()`-style
entry point. See the [porting status page](../../status.qmd) for where each of these lives.

## Density, CDF, and quantile

Two families illustrate the range hecfda covers: `Normal`, a light-tailed, symmetric distribution,
and `LogPearsonIII`, the heavy-tailed flood-frequency distribution HEC-FDA uses for annual peak
flows. Both plots below evaluate `dist_pdf()`, `dist_cdf()`, and `dist_quantile()` over a grid and
draw the three curves in the same earth-tone palette used throughout these examples.

```{r}
#| fig-cap: "Normal(mean = 0, sd = 1)"
x = seq(-4, 4, length.out = 200)
p = seq(0.001, 0.999, length.out = 200)
params = c(0, 1, 1)

par(mfrow = c(1, 3))
plot(x, dist_pdf("Normal", params, x), type = "l", col = "#6b705c", lwd = 2,
     xlab = "x", ylab = "density", main = "pdf")
plot(x, dist_cdf("Normal", params, x), type = "l", col = "#a5a58d", lwd = 2,
     xlab = "x", ylab = "probability", main = "cdf")
plot(p, dist_quantile("Normal", params, p), type = "l", col = "#cb997e", lwd = 2,
     xlab = "probability", ylab = "x", main = "quantile")
```

```{r}
#| fig-cap: "LogPearsonIII(mean = 3.3, sd = 0.254, skew = 0.0925, sample_size = 99)"
x = seq(100, 12000, length.out = 200)
p = seq(0.01, 0.99, length.out = 200)
params = c(3.3, 0.254, 0.0925, 99)

par(mfrow = c(1, 3))
plot(x, dist_pdf("LogPearsonIII", params, x), type = "l", col = "#6b705c", lwd = 2,
     xlab = "x", ylab = "density", main = "pdf")
plot(x, dist_cdf("LogPearsonIII", params, x), type = "l", col = "#a5a58d", lwd = 2,
     xlab = "x", ylab = "probability", main = "cdf")
plot(p, dist_quantile("LogPearsonIII", params, p), type = "l", col = "#cb997e", lwd = 2,
     xlab = "probability", ylab = "x", main = "quantile")
```

## Seeded sampling

`dist_sample()` draws `n` values by applying the seeded generator's uniform stream through
`inverse_cdf()`, the same path every Monte Carlo compute in hecfda uses. A histogram of 1000
seeded `Normal` draws should track the analytic density closely:

```{r}
#| fig-cap: "1000 seeded draws from Normal(0, 1) against the analytic pdf"
draws = dist_sample("Normal", c(0, 1, 1), n = 1000L, seed = 12345L)
hist(draws, breaks = 30, freq = FALSE, col = "#a5a58d", border = "white",
     xlab = "x", main = "")
x = seq(-4, 4, length.out = 200)
lines(x, dist_pdf("Normal", c(0, 1, 1), x), col = "#588157", lwd = 2)
```

`fda.dist_sample("Normal", [0, 1, 1], n=1000, seed=12345)` in Python draws the identical 1000
values, in the identical order, because both languages call the same seeded C++ generator.

## Faithful bugs

::: {.callout-warning}
This port reproduces upstream HEC-FDA behavior exactly, including a few known quirks in the C#
source. Two examples that affect distributions:

- `Empirical`'s pdf returns 0 for any `x` that is not an exact grid point in its data, rather than
  interpolating between points.
- `LogNormal` mixes natural-log (`ln`) and `log10` semantics inconsistently between its methods,
  matching the same inconsistency in the C# source rather than normalizing it.

These are not bugs in the port; they are the upstream behavior, transcribed on purpose so results
match HEC-FDA exactly. The [porting status page](../../status.qmd) describes what is ported and
what was left out.
:::
