---
title: "Getting started (R)"
---

hecfda ports the numerical core of USACE HEC-FDA to C++ with R and Python packages on top.
This example installs the package, draws from the seeded generator, and evaluates a
distribution. The draws below are identical in R, Python, and the upstream C#, because all
three run the same algorithm: .NET's seeded `System.Random`.

## Install

```{r}
#| eval: false
# install.packages("pak")
pak::pak("cameronbracken/hecfda/hecfdar")
```

## Seeded reproducibility

```{r}
library(hecfdar)
rng_sequence(seed = 1234L, n = 5L)
```

The same call in Python (`fda.rng_sequence(seed=1234, n=5)`) returns the same five numbers, and
so does `new Random(1234).NextDouble()` in .NET. Every Monte Carlo compute in hecfda draws from
this generator with fixed per-curve seeds, so full simulation results carry the same guarantee.

## A first distribution

```{r}
dist_cdf("Normal", c(0, 1, 1), 0) # 0.5
dist_quantile("LogPearsonIII", c(3.3, 0.254, 0.0925, 99), 1 - 0.01) # the "100-year" quantile
```

## A first EAD

The smallest useful compute: a uniform flow frequency, a linear flow-stage transform, one
stage-damage curve, one deterministic pass. This reproduces the HEC-FDA deterministic oracle of
150000 (the mean of the 600000-damage plateau over the frequency curve).

```{r}
res = ead_simulation(
  stage_damage = list(list(
    x = c(0, 15, 30), dist = "Uniform",
    params = list(c(0, 0, 10), c(0, 600000, 10), c(0, 600000, 10)),
    damage_category = "residential", asset_category = "Structure"
  )),
  flow_frequency = list(dist = "Uniform", params = c(0, 100000, 1000)),
  flow_stage = list(x = c(0, 100000), dist = "Uniform",
                    params = list(c(0, 0, 10), c(0, 30, 10))),
  min_iterations = 1L, max_iterations = 1L, deterministic = TRUE
)
res$total_ead
```
